### 2023-02-27 calculus.tex
Rewrite marked lambda caclulus section as discussed

\section{Motivating Examples}
\emph{Bidirectional typing} combines type checking, which determines if a program satisfies a given
type, and type synthesis, which generates a type from the program. This gives a simple algorithmic
system in which type information is recursively propagated through terms and allows for local type
inference.

@@

Often, programmers are working with program states that are syntactically well-formed but not
well-typed. However, formal typing semantics usually only give meaning to well-typed programs,
leaving ill-typed ones meaningless. This is insufficient for editor services and language servers,
which give users feedback on where and why a program is ill-typed. Traditionally, this has been
performed in an \emph{ad hoc} manner [citations?].

In this section, we motivate a formal calculus for bidirectional type error localization by
exploring different kinds of type errors that may occur.

% TODO state goals: prefer fine-grained over coarser blaming/marking

\subsection{Unbound Variables}
A simple and common example of an error that may occur in programs is the use of an unbound
variable. In this case, editor services often indicate visually that such usages are invalid.

% TODO image of an editor for this example? not sure what else to say about this

\subsection{Inconsistent Types}
Often, the type an expression \emph{should} be is known. Both operands of an addition expression
$\EPlus{\ETrue}{\ENumMV}$, for example, should have the type $\TNum$. In a bidirectional system,
$\ETrue$ and $\ENumMV$ would be analysed against $\TNum$, in which the former check would fail:
$\ETrue$ synthesises $\TBool$, which is inconsistent with $\TNum$.

% An editor may indicate such an error in different ways. Two obvious ones are considered below: (1)
% indicating an error on $\ETrue$, or (2) indicating an error on the entire expression
% $\EPlus{\ETrue}{\ENumMV}$.
% \[%
  % \EPlus{\ETrue}{\ENumMV} \qquad \EPlus{\ETrue}{\ENumMV}
% \]%
% The former ``blames'' the error on the operand,

% Since the ``fault'' of the error

% TODO cite real editor/language?

Since expressions typed in analytic mode are checked to satisfy a correct, known type, they are
given ``blame'' when ill-typed.

\subsection{Inconsistent Branches}

The previous section surveyed different type error cases. Now, we formally introduce a \emph{marking
calculus} for bidirectional type error localization on the gradually typed lambda calculus above.
This comes as a two-layer system:
%
\begin{itemize}
  \item \textbf{Unmarked expressions} (\cref{sec:calculus-uexp}), which compose the surface-level
    syntax. Types classify unmarked expressions according to a standard bidirectional gradual typing
    semantics.

  \item \textbf{Marked expressions} (\cref{sec:calculus-mexp}), which mirror the structure of
    unmarked expressions but are extended with \textbf{error marks}. Every marked expression
    corresponds to an unmarked expression by \emph{mark erasure}.
    % TODO more intuition for what marked expressions are?

  \item \textbf{Marking} (\cref{sec:calculus-marking}), which \textbf{marks} any unmarked expression into a marked
    expression, inserting error markings where appropriate.
\end{itemize}
%
That is, we extend the ordinary typing semantics of a language with a secondary ``marked'' language
and define a mapping from ``unmarked'' programs to ``marked'' programs. Note that while the
semantics that follow concern a gradually typed language, a non-gradual system may be recovered
straightforwardly by removing the unknown type and replacing type consistency with equality.

Key to ensuring correctness of the system---and, indeed, any such error localization semantics---is
a metatheorem, \emph{marking totality}, which guides the construction of rules:
%
\begin{theorem}[name=Marking Totality]
  Any syntactically well-formed unmarked expression may be marked into a marked expression.
\end{theorem}
%
We will give a more concrete statement of the metatheorem after formally introducing the components
of the system.

In \cref{sec:calculus-products}, we follow with an extension of the calculus to support binary
product types. In the interest of space, this overview focuses on ``interesting'' cases. All rules
and theorems may be found in the supplementary appendix, alongside a complete mechanization
(including the extension with binary products) in the Agda proof assistant, which is discussed in
\cref{sec:calculus-agda}.

% TODO extension to non-typing semantics?

\subsection{Unmarked Expressions}
\label{sec:calculus-uexp}

We consider a gradually typed lambda calculus extended with numbers and conditionals, whose syntax
and bidirectional typing are given in \cref{fig:calculus-syntax-example} and
\cref{fig:calculus-typing-example}, respectively. The base types $\TNum$ and $\TBool$ classify
number and boolean expressions. The number literal corresponding to the mathematical number $n$ is
given by $\ENumMV$, and there is a single addition operation on numeric expressions. $\ETrue$ and
$\EFalse$ correspond to the boolean values $\textsf{true}$ and $\textsf{false}$, and
$\EIf{\EMV_1}{\EMV_2}{\EMV_3}$ gives an if-else structure. In addition to these forms, $\TUnknown$
indicates an unknown type.

@@

Bidirectional typing employs two mutually defined judgments. \emph{Type synthesis}, written
$\ctxSynType{\ctx}{\EMV}{\TMV}$, establishes that, under the typing context $\ctx$, the expression
$\EMV$ synthesises or infers the type $\TMV$. \emph{Type analysis}, written
$\ctxAnaType{\ctx}{\EMV}{\TMV}$, states that the expression $\EMV$ may appear where an expression of
type $\TMV$ is expected.

The \emph{subsumption} rule (ASubsume) states that if an expression synthesises a type, it may be
analysed against that type or any \emph{consistent} type. To invoke subsumption, the expression
$\EMV$ must be ``subsumable'', written $\subsumable{\EMV}$, a judgment defined on all forms except
lambda abstractions and if-else expressions. This restricts the usage of subsumption to a rule of
``last resort'', admitting the unicity of typing.

% \begin{theorem}[name=Unmarked Expression Typing Unicity] \
  % If $\ctxSynType{\ctx}{\EMV}{\TMV}$ and $\ctxSynType{\ctx}{\EMV}{\TMV'}$, then $\TMV = \TMV'$.
% \end{theorem}

\input{figs/calculus-typing-uexp}

The \emph{consistency} judgment $\consistent{\TMV}{\TMV'}$ that appears in the subsumption rule
defines a reflexive and symmetric (but not transitive) relation between types. The unknown type is
additionally consistent with any other type, and arrow types are consistent if their components are.
The USAp and UALam rules for application and lambda abstractions are nearly standard, but employ the
judgment $\matchedArrow{\TMV}{\TMV_1}{\TMV_2}$, which gives that $\TMV$ is a \emph{matched arrow
type}, i.e. it may be considered an arrow type. It is purely a technical mechanism to avoid
duplication of rules related to arrow types.

\input{figs/calculus-type-judgments}

The branches of if-else expressions may synthesize different types $\TMV_1$ and $\TMV_2$ as long as
they are consistent. The entire expression then synthesizes the \emph{join}
$\TJoin{\TMV_1}{\TMV_2}$.

\subsection{Marked Expressions}
\label{sec:calculus-mexp}

Marked expressions, $\ECMV$, extend the syntax of unmarked expressions with \emph{explicit error
marks}. As given in the syntax definition of \cref{fig:calculus-syntax-mexp}, there are three such
forms, which mirror the examples discussed previously:
%
\begin{itemize} % TODO not the snappiest names
  \item $\ECUnbound{x}$, the \emph{unbound variable mark}, denotes the usage of an unbound variable.

  \item $\ECInconType{\ECMV}$, the \emph{inconsistent types mark}, occurs when the synthesised type
    of $\ECMV$ is inconsistent with the required type.

  \item $\ECInconBr{\ECMV_1}{\ECMV_2}{\ECMV_3}$, the \emph{inconsistent branches mark}, corresponds
    to an if-else expression where the types of the two branches $\ECMV_2$ and $\ECMV_3$ are
    inconsistent.
\end{itemize}
%
Additional forms may arise as required by the language.

% TODO examples of such?

\begin{figure}[htbp]
  \[\begin{array}{rrcl}
    \ECMName & \ECMV & \Coloneqq & x \mid \ECLam{x}{\TMV}{\ECMV} \mid \ECAp{\ECMV}{\ECMV} \mid \ECLet{x}{\ECMV}{\ECMV} \\
             &       & \mid         & \ECNumMV \mid \ECPlus{\ECMV}{\ECMV}
                       \mid           \ECTrue \mid \ECFalse \mid \ECIf{\ECMV}{\ECMV}{\ECMV} \\
             &       & \mid         & \ECUnbound{x} \mid \ECInconType{\ECMV} \mid \ECInconBr{\ECMV}{\ECMV}{\ECMV}
  \end{array}\]
  \caption{Syntax of marked expressions}
  \label{fig:calculus-syntax-mexp}
\end{figure}

The unmarked version of a marked expression $\ECMV$ may be recovered through \emph{mark erasure},
given by $\erase{\ECMV}$. As in \Cref{fig:calculus-mark-erasure}, this total function may be defined
inductively in a straightforward manner. Observe that erasing the inconsistent branches mark gives
an if-else expression with sub-expression marks recursively erased.

\input{figs/calculus-mark-erasure}

Marked expressions are also typed by a bidirectional system, which, for most forms, closely mirrors
the rules for corresponding unmarked forms. Additional rules handle the cases when certain premises
are not met, requiring then that expressions be wrapped in appropriate error marks.
\Cref{fig:calculus-typing-mexp} contains a selection of these ``interesting'' cases.

In parallel with the rule governing bound variables, MSUnbound types unbound variables. Such
variables must be marked unbound, and since nothing may be said about their types, they synthesise
the unknown type. MSInconsistentBranches corresponds to the USIf, handling the case that the
branches synthesise inconsistent types and hence have no join. Such forms are required to be marked
as such.

\input{figs/calculus-typing-mexp}

% TODO Not really sure how to explain the syntax and typing of marked expressions in an intuitive
% way in relation to unmarked expressions.

In addition to a synthesis rule for lambda abstractions that mirrors USLam, observe that there are
three, instead of one, analytic lambda rules. MALam1 applies to $\ECLam{x}{\TMV}{\ECMV}$ if the type
$\TMV_3$ analysed against is a matched arrow type $\TArrow{\TMV_1}{\TMV_2}$ and the input type
$\TMV_1$ is consistent with the binding ascription $\TMV$. This corresponds to UALam.

However, we are additionally concerned with the cases when $\TMV_3$ is not a matched arrow type or
$\TMV_1$ is inconsistent with $\TMV$. These are described by MALam2 and MALam3. The absence of a
matched arrow type means that $\TMV_3$ must be inconsistent the type synthesised by the lambda, so
the former appropriately requires that the lambda be inside an inconsistent types mark. In this
case, as no output type is known, the body is analysed against the unknown type. Similarly,
inconsistency between the matched input type and binding ascription also implies that $\TMV_3$ is
inconsistent with the synthesised type; MALam3 also requires that the lambda be marked inconsistent,
and the body is analysed against $\TMV_2$.

% TODO mention that every expression synthesises?

Two synthesis rules govern application. In the unmarked language, the application
$\EAp{\EMV_1}{\EMV_2}$ is well-typed if the type $\TMV$ synthesised by $\EMV_1$ is a matched arrow
type $\TArrow{\TMV_1}{\TMV_2}$ and the argument $\EMV_2$ analyses against $\TMV_1$. This corresponds
to the MSAp1 rule for marked expressions. In the case that $\TMV$ is not a matched arrow type, i.e.
it may not be applied, we require $\ECMV_1$ to be inside an inconsistent types mark. In this way, the
blame is placed upon $\ECMV_1$, and this approach is in accordance with the meaning of the
inconsistent types mark.

It is worth noting, however, that $\ECMV_1$ is not actually in analytic position but in synthetic
position. Treating such ``constrained synthetic'' mode specially, we may instead designate a
specialized mark, written $\ECInconMatchedArrow{\ECMV_1}$, for such applications. This yields the
following rule:
\[%
  \judgment{
    \ctxSynType{\ctx}{\ECMV_1}{\TMV} \\
    \notMatchedArrow{\TMV} \\
    \ctxAnaType{\ctx}{\ECMV_2}{\TUnknown}
  }{
    \ctxSynType{\ctx}{\ECApNonMatchedAltAlt{\ECMV_1}{\ECMV_2}}{\TUnknown}
  }{MSAp2'}
\]%
It then becomes natural to extend this to other elimination forms that require the handling of
unmatched types, such as products. We demonstrate such an extension in \cref{sec:calculus-products}.

Finally, alongside the standard subsumption rule, MAInconsistentTypes governs most occurrences of
the inconsistent types mark $\ECInconType{\ECMV}$ analyses against any type as long as $\ECMV$
synthesises an inconsistent type. In both cases, we again stipulate that subsumption may be applied
to $\ECMV$, written $\subsumable{\ECMV}$, for similar reasons as before.

\subsection{Marking}
\label{sec:calculus-marking}

The central component of the calculus is \emph{marking}, a bidirectional system given by the
following mutually defined judgments. The synthetic insertion judgment
$\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$ establishes that, under the context $\Gamma$, the
unmarked expression $\EMV$ is ``marked into'' the marked expression $\ECMV$, which synthesizes type
$\TMV$. Analogously, the analytic insertion judgment $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$
states that the unmarked expression $\EMV$ is marked into $\ECMV$, which analyses against type
$\TMV$.

% TODO need to better explain why it is bidirectional

\input{figs/calculus-mark-insertion}

The rules for mark insertion are given in \cref{fig:calculus-mark-insertion}. Observe that they
mirror the typing rules for marked expressions closely: all possible outcomes for fallible premises
are handled and markings are inserted as needed. For example, ISVar and ISUnbound together handle
variable usages. The former yields the same variable in the marked language when it is bound. The
latter marks the variable as unbound if it is not found in the context.

We again find two rules governing the marking of function application. ISAp1 corresponds to MSAp1 as
the ``success case'', recursively marking $\EMV_1$ and $\EMV_2$. Accordingly, ISAp2 parallels MSAp2
as the ``error case''. Similarly, ISIf and ISInconsistentBranches correspond to MSIf and
MSInconsistentBranches, and IALam1, IALam2, and IALam3 correspond to MSLam1, MSLam2, and MSLam3.
The rules IAInconsistentTypes and IASubsume, of course, act in parallel to MAInconsistentTypes and
MASubsume.

We may now formally restate the critical metatheorem of marking totality:
%
\begin{theorem}[name=Marking Totality] \
  \label{thm:calculus-marking-totality}
  \begin{enumerate}
    \item For all $\ctx$ and $\EMV$, there exist $\ECMV$ and $\TMV$ such that
      $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$.
  \item For all $\ctx$, $\EMV$, and $\TMV$, there exists $\ECMV$ such that
    $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$.
  \end{enumerate}
\end{theorem}

Additionally, it is important that marking preserves the original syntactic structure modulo error
marks; it would be trivial to satisfy totality by marking all expressions into numbers, for example.
This property is ensured by the following well-formedness metatheorem:
%
\begin{theorem}[name=Marking Well-Formedness] \
  \label{thm:calculus-marking-well-formedness}
  \begin{enumerate}
    \item If $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$, then $\ctxSynType{\ctx}{\ECMV}{\TMV}$ and
      $\erasesTo{\ECMV}{\EMV}$.
    \item If $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$, then $\ctxAnaType{\ctx}{\ECMV}{\TMV}$ and
      $\erasesTo{\ECMV}{\EMV}$.
  \end{enumerate}
  % TODO split this into two theorems?
  % TODO not really sure how to intuitively describe the part about types matching up
\end{theorem}
%
In other words, if an unmarked expression is marked into some marked expression, erasing error marks
from the marked expression gives the original unmarked expression. To ensure that we place error
marks in the correct spots, we furthermore require the following theorem:
%
\begin{theorem}[name=Marking of Well-Typed/Ill-Typed Expressions] \ % TODO better name for this
  \label{thm:calculus-marking-well-ill-typed}
  \begin{enumerate}
    \item \begin{enumerate}
        \item If $\ctxSynType{\ctx}{\EMV}{\TMV}$, then $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$
          such that $\imageIs{\EMV}{\ECMV}$.
        \item If $\ctxAnaType{\ctx}{\EMV}{\TMV}$, then $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$
          such that $\imageIs{\EMV}{\ECMV}$.
      \end{enumerate}

    \item \begin{enumerate}
        \item If there does not exist $\TMV$ such that $\ctxSynType{\ctx}{\EMV}{\TMV}$, then for all
          $\ECMV$ and $\TMV'$ such that $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV'}$, we have
          $\imageIsNot{\EMV}{\ECMV}$.
        \item If there does not exist $\TMV$ such that $\ctxAnaType{\ctx}{\EMV}{\TMV}$, then for all
          $\ECMV$ and $\TMV'$ such that $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV'}$, we have
          $\imageIsNot{\EMV}{\ECMV}$.
      \end{enumerate}
  \end{enumerate}
\end{theorem}
%
That is, since error marks indicate type errors in the program, a well-typed program should, of
course, have no error marks. The converse---that marks are inserted into ill-typed programs---is
also true. We have chosen to formulate the absence of error marks through an inclusion mapping of
unmarked expressions into marked expressions; alternatively, a predicate over marked expressions
could be used. % TODO latter is easier to do in Agda, but maybe it is weaker?

Finally, no less importantly, marking is deterministic. This is given by
\cref{thm:calculus-marking-unicity}.
%
\begin{theorem}[name=Marking Unicity] \
  \label{thm:calculus-marking-unicity}
  \begin{enumerate}
    \item If $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV_1}{\TMV_1}$ and
      $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV_2}{\TMV_2}$, then $\ECMV_1 = \ECMV_2$ and $\TMV_1 =
      \TMV_2$.
    \item If $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV_1}{\TMV_1}$ and
      $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV_2}{\TMV_2}$, then $\ECMV_1 = \ECMV_2$ and $\TMV_1 =
      \TMV_2$.
  \end{enumerate}
\end{theorem}
%
Together with totality, marking may be considered and implemented as a total function.

>>>>

Programmers often work with program states that are syntactically well-formed but not well-typed.
\Cref{fig:calculus-examples} shows common type errors as they appear in the Hazel programming
environment. For example, a simple error that may occur is the use of an unbound variable. As in
\cref{fig:calculus-examples-unbound}, editor services indicate the usage of unbound $y$ is invalid
by highlighting it in red.

\input{figs/calculus-examples}

Sometimes, the type an expression \emph{should} be is known. Consider
\cref{fig:calculus-examples-inconsistent-types}, in which it is expected that both operands of the
$+$ operator are numbers. Since $\textsf{true}$ is not a number, a type error arises, and Hazel
indicates a mismatch between the expected and actual types of the operand. The application in
\cref{fig:calculus-examples-app-non-lambda} of a non-lambda leads to a similar kind of error. In
this case, however, the expectation is not any singular type, but any member of a family of function
types. \Cref{fig:calculus-examples-inconsistent-branches} presents an error of inconsistent
branches: the branches of the if-else expression are of different types. Hazel opts to highlight the
entire expression in such a situation, while other tools may choose one branch to be ``correct'' and
indicate a type mismatch error on the other.

% TODO elaborate on this part

Unfortunately, formal typing semantics usually only give meaning to well-typed programs, leaving
such ill-typed ones meaningless. This is insufficient for editor services and language servers,
which must provide users feedback on where and why a program is ill-typed. Moreover, they must
continue providing editor services even when an error is encountered. Indeed, similarly to
error-recovering parsers, real-world type checkers must recover from type errors. Traditionally,
this has been performed in an \emph{ad hoc} manner [citations?], with different languages and tools
adopting varying approaches in the localization, reportage, and recovery from errors.

The key contribution of this section is a formal calculus, called the \emph{marking calculus}, for
type error localization in bidirectional systems, a judgmental framework provides a unified theory
for the design of typing semantics that supports ill-typed programs. \Cref{sec:calculus-calculus}
describes its form and metatheory by example on a minimal gradually typed lambda calculus, extending
with binary product types in \cref{sec:calculus-products}. All rules and theorems may be found in
the supplementary appendix, alongside a complete mechanization in the Agda proof assistant, which is
discussed in \cref{sec:calculus-agda}. Finally, in \cref{sec:calculus-poly}, we consider how the
system might be extended to more complex typing features, such as System F-style polymorphism.

% TODO not really sure when to start talking about bidirectional typing

\emph{Bidirectional typing} combines type checking, which determines if a program satisfies a given
type, and type synthesis, which generates a type from the program. This gives a simple algorithmic
system in which type information is recursively propagated through terms and allows for local type
inference.

@@

\subsection{The Calculus}
\label{sec:calculus-calculus}

% TODO outline the structure of this section at the beginning
% TODO make it clear unmarked/marked languages share same types

We now formally introduce the \emph{marking calculus} through its application to a gradually typed
lambda calculus extended with numbers and booleans. Consider the syntax for such a language, as
shown in \cref{fig:calculus-syntax-uexp}. The base types $\TNum$ and $\TBool$ classify number and
boolean expressions. The number literal corresponding to the mathematical number $n$ is given by
$\ENumMV$, and there is a single addition operation on numeric expression. $\ETrue$ and $\EFalse$
correspond to the boolean values $\textsf{true}$ and $\textsf{false}$, respectively, and
$\EIf{\EMV_1}{\EMV_2}{\EMV_3}$ gives an if-else structure. In addition to these forms, $\TUnknown$
is the unknown type from gradual type theory.

@@

The language is typed bidirectionally, which employs two mutually defined judgments. \emph{Type
synthesis}, written $\ctxSynType{\ctx}{\EMV}{\TMV}$, establishes that, under the typing context
$\ctx$, the expression $e$ synthesises or infers the type $\TMV$. \emph{Type analysis}, written
$\ctxAnaType{\ctx}{\EMV}{\TMV}$, states that the expression $\EMV$ may appear where an expression of
type $\TMV$ is expected. We may give these typing semantics traditionally, without considering error
localization and recovery.

The \emph{marking calculus} builds atop these typing semantics into a two-layer system:
%
\begin{itemize}
  \item The \textbf{unmarked language}, which is the original language. Expressions of this language
    are \emph{unmarked expressions}.

  \item The \textbf{marked language}, which mirrors the structure of the unmarked language but is
    extended with explicit \emph{error marks}. We call expressions of this language to be
    \emph{marked expressions}, denoted $\ECMV$.
    % TODO more intuition for what marked expressions are?

  \item \textbf{Marking}, which \emph{marks} any unmarked expression into a marked expression,
    inserting error markings where appropriate.
\end{itemize}
%
In other words, we extend the ordinary typing semantics of a language with a secondary ``marked''
language and ``mark'' unmarked programs into marked ones. This corresponds to a type checking
process with error localization and recovery.

% TODO more intuition

We give, for now, only part of the marked expression syntax in \cref{fig:calculus-syntax-mexp-part}.
Throughout the remainder of this section, as we consider the formulation of marking, we will add the
necessary error mark forms.

% TODO include holes
\begin{figure}[htbp]
  \[\begin{array}{rrcl}
    \ECMName & \ECMV & \Coloneqq & x \mid \ECLam{x}{\TMV}{\ECMV} \mid \ECAp{\ECMV}{\ECMV} \mid \ECLet{x}{\ECMV}{\ECMV} \\
             &       & \mid         & \ECNumMV \mid \ECPlus{\ECMV}{\ECMV}
                       \mid           \ECTrue \mid \ECFalse \mid \ECIf{\ECMV}{\ECMV}{\ECMV} \\
             &       & \mid         & \cdots
             % &       & \mid         & \ECUnbound{x} \mid \ECInconType{\ECMV} \mid \ECInconBr{\ECMV}{\ECMV}{\ECMV}
  \end{array}\]
  \caption{Partial syntax of marked expressions}
  \label{fig:calculus-syntax-mexp-part}
\end{figure}

The marked language possesses its own typing semantics, also formulated bidirectionally. We write
$\ctxSynType{\ctx}{\ECMV}{\TMV}$ for synthesis and $\ctxAnaType{\ctx}{\ECMV}{\TMV}$ for analysis.
Finally, marking is bidirectional as well. The synthetic marking judgment
$\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$ establishes that, under the context $\Gamma$, the
unmarked expression $\EMV$ is ``marked into'' the marked expression $\ECMV$, which synthesizes type
$\TMV$. Analogously, the analytic marking judgment $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$
states that $\EMV$ is marked into $\ECMV$, which analyses against $\TMV$.

How can we ensure that the marking procedure is correctly defined? There are two critical
metatheorems to keep in mind as we continue. The first is a \emph{totality} of marking:
%
\begin{theorem}[name=Marking Totality] \
  \label{thm:calculus-marking-totality}
  \begin{enumerate}
    \item For all $\ctx$ and $\EMV$, there exist $\ECMV$ and $\TMV$ such that
      $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$.
  \item For all $\ctx$, $\EMV$, and $\TMV$, there exists $\ECMV$ such that
    $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$.
  \end{enumerate}
\end{theorem}
%
That is, we may mark, i.e. type check and localize errors for \emph{any} syntactically well-formed
program. Furthermore, marking should preserve the syntactic structure of programs, modulo error
marks. Thus, we must define a notion of \emph{mark erasure}, which converts marked expressions back
into unmarked expression by removing error marks. \Cref{fig:calculus-mark-erasure-part} gives the
definition of mark erasure, written $\erase{\ECMV}$, for our initial marked expression syntax. Then,
we may define a \emph{well-formedness} criterion for marking:
%
\begin{theorem}[name=Marking Well-Formedness] \
  \label{thm:calculus-marking-well-formedness}
  \begin{enumerate}
    \item If $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$, then $\ctxSynType{\ctx}{\ECMV}{\TMV}$ and
      $\erasesTo{\ECMV}{\EMV}$.
    \item If $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$, then $\ctxAnaType{\ctx}{\ECMV}{\TMV}$ and
      $\erasesTo{\ECMV}{\EMV}$.
  \end{enumerate}
  % TODO split this into two theorems?
  % TODO not really sure how to intuitively describe the part about types matching up
\end{theorem}
%
In other words, if an unmarked expression is marked into some marked expression, erasing error marks
from the marked expression gives the original unmarked expression. We also ensure that marking gives
marked expressions that are well-typed; indeed, their types should match the types synthesised or
analysed against in marking. % TODO more intuition for this theorem

\input{figs/calculus-mark-erasure-part}

In the remainder of this subsection, we consider the different type errors that may arise in the
unmarked language; in this way, the typing semantics of the unmarked language guide the construction
of the marking process.

\subsubsection{Variables}
\label{sec:calculus-variables}

% TODO subsumption is discussed later...
The typing of unmarked variables in the gradually typed lambda calculus is standard. Because of
subsumption, we need only define a synthesis rule:
\[%
  \judgment{
    \inCtx{\ctx}{x}{\TMV}
  }{
    \ctxSynType{\ctx}{x}{\TMV}
  }{USVar}
\]%
If $x$ is bound to a type in the typing context, it synthesizes that type. Straightforwardly,
marking would convert the unmarked $x$ into the same marked $x$ via the following marking and typing
rules:
%
\begin{mathpar}
  \judgment{
    \inCtx{\ctx}{x}{\TMV}
  }{
    \ctxSynFixedInto{\ctx}{x}{x}{\TMV}
  }{ISVar}

  \judgment{
    \inCtx{\ctx}{x}{\TMV}
  }{
    \ctxSynType{\ctx}{x}{\TMV}
  }{MSVar}
\end{mathpar}
% TODO more intuition on why this makes sense?

But what happens a variable, such as $y$ in \cref{fig:calculus-examples-unbound}, is \emph{not}
bound? Then, \textsc{USVar} does not apply, and the typing semantics say nothing about such
expressions. To satisfy totality, such a possibility motivates an \emph{unbound variable mark}:
%
\[\begin{array}{rrcl}
  \ECMName & \ECMV & \Coloneqq & \cdots \mid \ECUnbound{x}
\end{array}\]
%
accompanied by the following rules:
%
\begin{mathpar}
  \judgment{
    \notInCtx{\ctx}{x}
  }{
    \ctxSynFixedInto{\ctx}{x}{\ECUnbound{x}}{\TUnknown}
  }{ISUnbound}

  \judgment{
    \notInCtx{\ctx}{x}
  }{
    \ctxSynType{\ctx}{\ECUnbound{x}}{\TUnknown}
  }{MSUnbound}
\end{mathpar}
%
\textsc{ISUnbound} states that an variable unbound in the context is marked into that variable
inside an unbound variable mark. \textsc{MSUnbound} then gives the typing of unbound variable marks.
Since nothing may be said about their types, we synthesize the unknown type, which allows type
checking to continue---instead of simply failing.

To prove \cref{thm:calculus-marking-well-formedness}, we extend the definition of mark erasure to
handle the new mark:
%
\[\begin{array}{rcl}
  & \cdots & \\
  \erasesToRow{\ECUnbound{x}}{x}
\end{array}\]

\subsubsection{Lambda Abstractions}
\label{sec:calculus-lambda-abstractions}

There are explicit synthesis and analysis rules for lambda abstractions:
%
\begin{mathpar}
  \judgment{
    \ctxSynType{\extendCtx{\ctx}{x}{\TMV}}{\EMV}{\TMV_2}
  }{
    \ctxSynType{\ctx}{\ELam{x}{\TMV_1}{\EMV}}{\TArrow{\TMV_1}{\TMV_2}}
  }{USLam}

  \judgment{
    \matchedArrow{\TMV_3}{\TMV_1}{\TMV_2} \\
    \consistent{\TMV}{\TMV_1} \\
    \ctxAnaType{\extendCtx{\ctx}{x}{\TMV}}{\EMV}{\TMV_2}
  }{
    \ctxAnaType{\ctx}{\ECLam{x}{\TMV}{\EMV}}{\TMV_3}
  }{UALam}
\end{mathpar}
%
The synthesis rule is standard. In the premises of \textsc{UALam}, we use the notion of
\emph{type consistency} from gradual type theory, which defines a reflexive and symmetric (but not
transitive) relation between types, writing $\consistent{\TMV_1}{\TMV_2}$ to mean that $\TMV_1$ is
consistent with $\TMV_2$. This replaces the notion of \emph{type equality} to relate the unknown
type to all other types. The analysis rule also employs the judgment
$\matchedArrow{\TMV}{\TMV_1}{\TMV_2}$, which establishes that $\TMV$ is a \emph{matched arrow type},
i.e. it may be considered an arrow type. This notion is purely a technical mechanism to avoid
duplication of rules related to arrow types. \Cref{fig:calculus-consistency-matched-arrow} gives the
rules for both judgments.

% TODO mention explicitly that by inductive hypothesis, the marking of body cannot fail?
% TODO more intuition on these two judgments
% TODO more intuition on USLam and UALam

\begin{figure}[htbp]
  \raggedright
  \judgbox{\ensuremath{\consistent{\TMV_1}{\TMV_2}}} $\TMV_1$ is consistent with $\TMV_2$
  %
  \begin{mathpar}
    \judgment{ }{
      \consistent{\TUnknown}{\TMV}
    }{TCUnknown1}

    \judgment{ }{
      \consistent{\TMV}{\TUnknown}
    }{TCUnknown2}

    \judgment{ }{
      \consistent{\TMV}{\TMV}
    }{TCRefl}

    \judgment{
      \consistent{\TMV_1}{\TMV_1'} \\
      \consistent{\TMV_2}{\TMV_2'} \\
    }{
      \consistent{\TArrow{\TMV_1}{\TMV_2}}{\TArrow{\TMV_1'}{\TMV_2'}}
    }{TCArr} \\
  \end{mathpar}

  \judgbox{\ensuremath{\matchedArrow{\TMV}{\TMV_1}{\TMV_2}}} $\TMV$ has matched arrow type $\TArrow{\TMV_1}{\TMV_2}$
  %
  \begin{mathpar}
    \judgment{ }{
      \matchedArrow{\TUnknown}{\TUnknown}{\TUnknown}
    }{TMAHole}

    \judgment{ }{
      \matchedArrow{\TArrow{\TMV}{\TMV}}{\TMV}{\TMV}
    }{TMAArr}
  \end{mathpar}
  %
  \caption{Type consistency and matched arrow types.}
  \label{fig:calculus-consistency-matched-arrow}
\end{figure}

The synthesis rule for marking intuitively follows \textsc{USLam} closely. We should recursively
mark the body with an extended context binding $x$ to $\TMV$ and construct a new marked lambda
abstraction:
%
\begin{mathpar}
  \judgment{
    \ctxSynFixedInto{\extendCtx{\ctx}{x}{\TMV_1}}{\EMV}{\ECMV}{\TMV_2}
  }{
    \ctxSynFixedInto{\ctx}{\ELam{x}{\TMV_1}{\EMV}}{\ELam{x}{\TMV_1}{\ECMV}}{\TArrow{\TMV_1}{\TMV_2}}
  }{ISLam}

  \judgment{
    \ctxSynType{\extendCtx{\ctx}{x}{\TMV}}{\ECMV}{\TMV_2}
  }{
    \ctxSynType{\ctx}{\ECLam{x}{\TMV_1}{\ECMV}}{\TArrow{\TMV_1}{\TMV_2}}
  }{MSLam}
\end{mathpar}
%
Similarly, we construct corresponding analytic rules:
%
\begin{mathpar}
  \judgment{
    \matchedArrow{\TMV_3}{\TMV_1}{\TMV_2} \\
    \consistent{\TMV}{\TMV_1} \\\\
    \ctxAnaFixedInto{\extendCtx{\ctx}{x}{\TMV}}{\EMV}{\ECMV}{\TMV_2}
  }{
    \ctxAnaFixedInto{\ctx}{\ELam{x}{\TMV}{\EMV}}{\ECLam{x}{\TMV}{\ECMV}}{\TMV_3}
  }{IALam1}
